An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-10-19 scagnostics.Rmd
application 2 cleaned up

Figure \ref{fig:bbhSimulation} shows a scatterplot matrix analogue to that shown in Figure \ref{fig:neutronStarSPLOM} but where we have removed a subset of less informative parameters to obtain a clear picture. Strong association can be seen between the parameters ra, dec and time, where the first two describe the position of the event in the sky, and time is the merging time (in GPS units). Moreover, the relation between dec and time is not functional, which motivates us to use the TIC index for this example (a drawback is however the run time, this data sample contains 10k points, making optimisation of the MINE index functions very inefficient). In general we observe non-standard distributions for all variables, as well as various patterns of association between sets of variables. These patterns are representative of the likelihood function, since most priors are flat, or build such to capture growth with volume rather than distance. This is still efficient thanks to special sampling techniques that allow jumps in parameter space.

### Data pre-pocessing
We follow the same steps of data pre-processing as described above for the first example. First, since constraints on the masses of two systems observed via gravitational waves are strongly correltated, we again remove the smaller mass m2 from the set of parameters. We then rescale all parameters to the range from 0 to 1.

@@

As a first test we use the full dataset and compare results obtained with different index functions. Here we expect to recover the relation between the correlated variables dec, time and ra. We compare the final view using three different metrics in Figure \ref{fig:bbhGuided}, showing the views with (local?) maximum of the TIC index (left), spines2D index (middle) and for comparison the maximum of 1-convex (right). Indeed the TIC and spline index identify dec, time and ra parameters and show views of the 2-d band that is embedded in three dimensions \textcolor{red}{maybe having tour of those three dimensions would be helpful here}. Notably, the exact views differ based on the index function selected, optimising TIC is showing us the band "from the side". On the other hand, the index based on the scagnostics member convex is too noisy to find any views of interest, and the final  view is essentially equal to the starting plane in this case. All index values for the three final views are listed in Table \ref{tab:indexTableBBH}, where we see that indeed the splines2D index does not change between the first two views, while the TIC index is increased for the first view wrt the second. Finally the value of 1-convex is clearly larger for the first view as compared to the final view, but the optimisation did not succeed for this noisy index function.

We next follow a similar strategy as above and remove the variables with the strongest association from the dataset to understand if additional structure is encoded in the distribution. Removing thus time, dec and ra variable from the input, we optimise again the TIC, splines2d and 1-convex index functions. The results are shown in Fig. \ref{fig:bbhGuided2}. In this case the TIC index (left) identifies a view presenting seemingly linear correlation, originating however from one well constrained direction orthogonal to the main distribution of points. This direction is defined by the chi\_tot and chi\_p variables, for which we can read off the strong correlation in Fig. \ref{fig:bbhSimulation}. This direction is also found by the splines2D index (middle), which in addition identified the bimodal structure in the data stemming from the degeneracy between two plausible values for theta\_jn. (Interestingly the hole index in two dimensions does not identify this view, but is expected to find it in a 1-d guided tour.) Finally, as shown on the right, 1-convex does not identify any views of interest here.

We next investigate the dependence of the final result on the input starting plane. In particular if there are many local maxima the effect may be large, and many random starting planes may be needed to find a global maximum. To this end we first reduce the dataset dimensionality to include only two types of parameters: first, spin related parameters (i.e. alpha, theta\_jn, chi\_tot and chi\_p), and second position related parameters (i.e. ra, dec and distance). Note that this includes a number of variables for which we have established association and one dimensional structure above, i.e. ra and dec; chi\_tot and chi\_p; and theta\_jn. We generate a small set of random starting planes and optimise the splines2D index (it is prefered here over the TIC index as it is much more efficient to compute on large data sets). We compare six final views identified in this manner (i.e. differing only in randomly selected starting plane) in Fig. \ref{fig:bbhGuided3}. We note large variation in final views, only one of the six tries identified the view of ra vs dec, which however clearly has the maximum index value amongst all views. \textcolor{red}{add axes to plot, then we can make statement about how "far" the other views are from the "best" one.}

Finally, given the results above, it makes sense to also consider smaller parameter-subspaces to look for stronger association. (A procedure to first scout the parameter space for large index values is not efficent in this case because of the strong dependence on rotation angles in this paricular setup.) One subspace for which we found strong association between three variables is using the randomly selected parameter set (m1, ra, chi\_tot, alpha, distance, dec). Optimising again the splines2D index value allows us to find the final view shown in Figure \ref{constructedExample} (left) \textcolor{red}{(The small constribution of alpha is hidden in the axes here, middle plot is without rescaling, therefore same size contribution from dec an alpha to y axis)}, which we can reproduce from the original parameters as shown in the middle, as a combination of the position parameters ra and dec with the spin parameter alpha. For comparison we also show the dependence between ra and dec only, which is fuzzy compared to the ideal view identified by the PP optimisation. \textcolor{red}{This result is unexpected, it can be understood from ????? (Does it mean that knowing the precession angle gives more precise information about the location? Knowing ra uniquely determines dec-alpha rather than dec it seems.)}

```{r bbhGuided, results="hide", fig.cap="Final views identified by the guided tour using the TIC index (left), the splines2d index (middle) or 1-convex (right).", dev = "png", dpi=300}
bbhM <- rescale(select(bbhD,-m2, -chi_eff))
if(!file.exists("cache/bbhGuidedTIC.rda")){
  set.seed(2018)
  bbhTour1 <- save_history(bbhM, guided_tour(mineIndex("TIC")))
  bbhTourFull1 <- as.list(interpolate(bbhTour1))
  save(bbhTour1, bbhTourFull1, file = "cache/bbhGuidedTIC.rda")
} else {
  load("cache/bbhGuidedTIC.rda")
}
if(!file.exists("cache/bbhGuidedSpline.rda")){
  set.seed(2018)
  bbhTour2 <- save_history(bbhM, guided_tour(splineIndex()))
  bbhTourFull2 <- as.list(interpolate(bbhTour2))
  save(bbhTour2, bbhTourFull2, file = "cache/bbhGuidedSpline.rda")
} else {
  load("cache/bbhGuidedSpline.rda")
}
if(!file.exists("cache/bbhGuidedConvex.rda")){
  set.seed(2018)
  bbhTour3 <- save_history(bbhM, guided_tour(invConvexIndex()))
  bbhTourFull3 <- as.list(interpolate(bbhTour3))
  save(bbhTour3, bbhTourFull3, file = "cache/bbhGuidedConvex.rda")
} else {
  load("cache/bbhGuidedConvex.rda")
}

iLast <- length(bbhTourFull1)
fProj1 <- bbhTourFull1[[iLast]]
dProj1 <- as.tibble(bbhM %*% fProj1)
colnames(fProj1) <- c("PP1", "PP2")
colnames(dProj1) <- c("PP1", "PP2")
fProj1 <- as_tibble(fProj1) %>%
  add_column(label=colnames(bbhM))
p1 <- ggplot(dProj1, aes(PP1,PP2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1) + 
  geom_path(data=getCircle(0.4,0.5,0.26,0.24), aes(x=x, y=y)) + 
  xlim(0.1, 1.4) + ylim(-0.3, 0.9) +
  geom_segment(data=getAxes(fProj1, 0.4,0.5,0.26,0.24), aes(x=x1, xend=x2, y=y1, yend=y2)) +
  geom_text(data=getAxes(fProj1, 0.4,0.5,0.26,0.24), aes(x=x2, y=y2, label=label), size=2.5)
iLast <- length(bbhTourFull2)
fProj2 <- bbhTourFull2[[iLast]]
dProj2 <- as.tibble(bbhM %*% fProj2)
colnames(fProj2) <- c("PP1", "PP2")
colnames(dProj2) <- c("PP1", "PP2")
fProj2 <- as_tibble(fProj2) %>%
  add_column(label=colnames(bbhM))
p2 <- ggplot(dProj2, aes(PP1,PP2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1) + 
  geom_path(data=getCircle(0.3,-1,0.24,0.22), aes(x=x, y=y)) + 
  xlim(-0.1, 1.1) + ylim(-1.3, -0.2) +
  geom_segment(data=getAxes(fProj2,0.3,-1,0.24,0.22), aes(x=x1, xend=x2, y=y1, yend=y2)) +
  geom_text(data=getAxes(fProj2, 0.3,-1,0.24,0.22), aes(x=x2, y=y2, label=label), size=2.5)
iLast <- length(bbhTourFull3)
fProj3 <- bbhTourFull3[[iLast]]
dProj3 <- as.tibble(bbhM %*% fProj3)
colnames(fProj3) <- c("PP1", "PP2")
colnames(dProj3) <- c("PP1", "PP2")
fProj3 <- as_tibble(fProj3) %>%
  add_column(label=colnames(bbhM))
p3 <- ggplot(dProj3, aes(PP1,PP2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1) + 
  geom_path(data=getCircle(0.75, 1, 0.3, 0.28), aes(x=x, y=y)) + 
  xlim(-0.2, 1.3) + ylim(-0.1, 1.3) +
  geom_segment(data=getAxes(fProj3,0.75, 1, 0.3, 0.28), aes(x=x1, xend=x2, y=y1, yend=y2)) +
  geom_text(data=getAxes(fProj3, 0.75, 1, 0.3, 0.28), aes(x=x2, y=y2, label=label), size=2.5)
grid.arrange(p1, p2, p3, ncol=3) 
```

```{r indexTableBBH}
getIdxComp <- function(dProj, n){
  ticI <- mine(dProj$PP1, dProj$PP2)$TIC
  splineI <- splines2d(dProj$PP1, dProj$PP2)
  invConvI <- 1-scagnostics(as.matrix(dProj))[,"Convex"]
  return(tibble(TIC=ticI, splines2D=splineI, invConv=invConvI, nView=n))
}
idxT <- tibble(TIC=numeric(),
               splines2D=numeric(),
               invConv=numeric(),
               nView=numeric())
idxT <- rbind(idxT, getIdxComp(dProj1, 1),
      getIdxComp(dProj2, 2),
      getIdxComp(dProj3, 3))
colnames(idxT) <- c("TIC", "splines2D", "1-convex", "Projection")
knitr::kable(idxT,  caption = "Matrix of index values, for all three indices and all three final projections. \\label{tab:indexTableBBH}", digits=2) 
```

```{r bbhGuided2, results="hide", fig.cap="Final views identified by the guided tour using the TIC index (left), the splines2d index (middle) or 1-convex (right).", dev = "png", dpi=300}
bbhM2 <- rescale(select(bbhD,-m2, -chi_eff, -time, -dec, -ra))
if(!file.exists("cache/bbhGuidedTIC2.rda")){
  set.seed(2018)
  bbhTour12 <- save_history(bbhM2, guided_tour(mineIndex("TIC")))
  bbhTourFull12 <- as.list(interpolate(bbhTour12))
  save(bbhTour12, bbhTourFull12, file = "cache/bbhGuidedTIC2.rda")
} else {
  load("cache/bbhGuidedTIC2.rda")
}
if(!file.exists("cache/bbhGuidedSpline2.rda")){
  set.seed(2018)
  bbhTour22 <- save_history(bbhM2, guided_tour(splineIndex()))
  bbhTourFull22 <- as.list(interpolate(bbhTour22))
  save(bbhTour22, bbhTourFull22, file = "cache/bbhGuidedSpline2.rda")
} else {
  load("cache/bbhGuidedSpline2.rda")
}
if(!file.exists("cache/bbhGuidedConvex2.rda")){
  set.seed(2018)
  bbhTour32 <- save_history(bbhM2, guided_tour(invConvexIndex()))
  bbhTourFull32 <- as.list(interpolate(bbhTour32))
  save(bbhTour32, bbhTourFull32, file = "cache/bbhGuidedConvex2.rda")
} else {
  load("cache/bbhGuidedConvex2.rda")
}

@@

@@

@@

>>>>

This data contains posterior samples for simulation from a model describing a binary black hole (BBH) merger event. There are twelve model parameters. Flat priors are used for most model parameters.

Figure \ref{fig:bbhSimulation} shows a scatterplot matrix, of nine of the twelve parameters. (Parameter m2 is not shown because it is strongly linearly associated with m1, phi$\_$jl and psi are not shown because they uniform, and not associated with other parameters.) Among the nine plotted parameters, strong nonlinear relationships can be seen between the parameters ra, dec and time. The first two describe the position of the event in the sky, and time is the merging time (in GPS units). Because of the elliptical relationship between dec and time, the TIC index is used for PP, even though it is slow to compute. Between the other parameters, the main structure seen is multimodality and some skewness. These patterns are representative of the likelihood function, since most priors are flat, or built to capture growth with volume rather than distance.

### Data pre-processing

The analysis is conducted on 11 of the twelve parameters. One variable is removed, m2, because it is so strongly associated with m1. All parameters are scaled into the range 0 to 1.

@@

#### Exploring 11D with all PP indexes

All seven PP indexes are applied to the data. Figure \ref{fig:bbhGuided} showing the projections that maximise three of the indexes. TIC and splines2D indexes identify very similar projections, that are based on the three parameters, dec, time and ra. This is ot be expected based on the pairwise scatterplots. \textcolor{red}{maybe having tour of those three dimensions would be helpful here}. On the other hand, the 1-convex index finds a very different view, but actually the problem is that the optimisation doesn't adequately reach a maximum for this index.

```{r bbhGuided, results="hide", fig.cap="Projections corresponding to the maxima of three indices: TIC, splines2D and 1-convex. Projections a, b found by TIC and splines3D are very similar, and involve the same three parameters, ra, dec and time. The 1-convex index finds a very different view.", dev = "png", dpi=300}
bbhM <- rescale(select(bbhD,-m2, -chi_eff))
if(!file.exists("cache/bbhGuidedTIC.rda")){
  set.seed(2018)
  bbhTour1 <- save_history(bbhM, guided_tour(mineIndex("TIC")))
  bbhTourFull1 <- as.list(interpolate(bbhTour1))
  save(bbhTour1, bbhTourFull1, file = "cache/bbhGuidedTIC.rda")
} else {
  load("cache/bbhGuidedTIC.rda")
}
if(!file.exists("cache/bbhGuidedSpline.rda")){
  set.seed(2018)
  bbhTour2 <- save_history(bbhM, guided_tour(splineIndex()))
  bbhTourFull2 <- as.list(interpolate(bbhTour2))
  save(bbhTour2, bbhTourFull2, file = "cache/bbhGuidedSpline.rda")
} else {
  load("cache/bbhGuidedSpline.rda")
}
if(!file.exists("cache/bbhGuidedConvex.rda")){
  set.seed(2018)
  bbhTour3 <- save_history(bbhM, guided_tour(invConvexIndex()))
  bbhTourFull3 <- as.list(interpolate(bbhTour3))
  save(bbhTour3, bbhTourFull3, file = "cache/bbhGuidedConvex.rda")
} else {
  load("cache/bbhGuidedConvex.rda")
}

iLast <- length(bbhTourFull1)
fProj1 <- bbhTourFull1[[iLast]]
dProj1 <- as.tibble(bbhM %*% fProj1)
colnames(fProj1) <- c("PP1", "PP2")
colnames(dProj1) <- c("PP1", "PP2")
fProj1 <- as_tibble(fProj1) %>%
  add_column(label=colnames(bbhM))
p1 <- ggplot(dProj1, aes(PP1,PP2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1) + 
  geom_path(data=getCircle(0.4,0.5,0.26,0.24), aes(x=x, y=y)) + 
  xlim(0.1, 1.4) + ylim(-0.3, 0.9) +
  geom_segment(data=getAxes(fProj1, 0.4,0.5,0.26,0.24), aes(x=x1, xend=x2, y=y1, yend=y2)) +
  geom_text(data=getAxes(fProj1, 0.4,0.5,0.26,0.24), aes(x=x2, y=y2, label=label), size=2.5) + ggtitle("a. TIC: 779.46")
iLast <- length(bbhTourFull2)
fProj2 <- bbhTourFull2[[iLast]]
dProj2 <- as.tibble(bbhM %*% fProj2)
colnames(fProj2) <- c("PP1", "PP2")
colnames(dProj2) <- c("PP1", "PP2")
fProj2 <- as_tibble(fProj2) %>%
  add_column(label=colnames(bbhM))
p2 <- ggplot(dProj2, aes(PP1,PP2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1) + 
  geom_path(data=getCircle(0.3,-1,0.24,0.22), aes(x=x, y=y)) + 
  xlim(-0.1, 1.1) + ylim(-1.3, -0.2) +
  geom_segment(data=getAxes(fProj2,0.3,-1,0.24,0.22), aes(x=x1, xend=x2, y=y1, yend=y2)) +
  geom_text(data=getAxes(fProj2, 0.3,-1,0.24,0.22), aes(x=x2, y=y2, label=label), size=2.5) + ggtitle("b. splines2D: 0.97")
iLast <- length(bbhTourFull3)
fProj3 <- bbhTourFull3[[iLast]]
dProj3 <- as.tibble(bbhM %*% fProj3)
colnames(fProj3) <- c("PP1", "PP2")
colnames(dProj3) <- c("PP1", "PP2")
fProj3 <- as_tibble(fProj3) %>%
  add_column(label=colnames(bbhM))
p3 <- ggplot(dProj3, aes(PP1,PP2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1) + 
  geom_path(data=getCircle(0.75, 1, 0.3, 0.28), aes(x=x, y=y)) + 
  xlim(-0.2, 1.3) + ylim(-0.1, 1.3) +
  geom_segment(data=getAxes(fProj3,0.75, 1, 0.3, 0.28), aes(x=x1, xend=x2, y=y1, yend=y2)) +
  geom_text(data=getAxes(fProj3, 0.75, 1, 0.3, 0.28), aes(x=x2, y=y2, label=label), size=2.5) + ggtitle("c. 1-convex: 0.50")
grid.arrange(p1, p2, p3, ncol=3) 
```

```{r indexTableBBH, eval=FALSE}
getIdxComp <- function(dProj, n){
  ticI <- mine(dProj$PP1, dProj$PP2)$TIC
  splineI <- splines2d(dProj$PP1, dProj$PP2)
  invConvI <- 1-scagnostics(as.matrix(dProj))[,"Convex"]
  return(tibble(TIC=ticI, splines2D=splineI, invConv=invConvI, nView=n))
}
idxT <- tibble(TIC=numeric(),
               splines2D=numeric(),
               invConv=numeric(),
               nView=numeric())
idxT <- rbind(idxT, getIdxComp(dProj1, 1),
      getIdxComp(dProj2, 2),
      getIdxComp(dProj3, 3))
colnames(idxT) <- c("TIC", "splines2D", "1-convex", "Projection")
knitr::kable(idxT,  caption = "Matrix of index values, for all three indices and all three final projections. \\label{tab:indexTableBBH}", digits=2) 
```

#### Exploring reduced space

The variables time, dec and ra are dropped from the data, and PP is applied to the remaining 8D space. Figure \ref{fig:bbhGuided2} shows the projections which maximise the TIC, splines2D and 1-convex indices. The results provide similar information as laready learned from the scatterplot matrix (Figure \ref{bbhSimulation}). The parameters chi$\_$tot and chi$\_$p are linearly related (TIC maxima), and theta$\_$jn has a bimodal distribution yielding the figure 8 shape found by the splines2D index. The 1-convex index finds nothing interesting.

```{r bbhGuided2, results="hide", fig.cap="Projections of the reduced 8D space corresponding to the maxima of three indices: TIC, splines2D and 1-convex.", dev = "png", dpi=300}
bbhM2 <- rescale(select(bbhD,-m2, -chi_eff, -time, -dec, -ra))
if(!file.exists("cache/bbhGuidedTIC2.rda")){
  set.seed(2018)
  bbhTour12 <- save_history(bbhM2, guided_tour(mineIndex("TIC")))
  bbhTourFull12 <- as.list(interpolate(bbhTour12))
  save(bbhTour12, bbhTourFull12, file = "cache/bbhGuidedTIC2.rda")
} else {
  load("cache/bbhGuidedTIC2.rda")
}
if(!file.exists("cache/bbhGuidedSpline2.rda")){
  set.seed(2018)
  bbhTour22 <- save_history(bbhM2, guided_tour(splineIndex()))
  bbhTourFull22 <- as.list(interpolate(bbhTour22))
  save(bbhTour22, bbhTourFull22, file = "cache/bbhGuidedSpline2.rda")
} else {
  load("cache/bbhGuidedSpline2.rda")
}
if(!file.exists("cache/bbhGuidedConvex2.rda")){
  set.seed(2018)
  bbhTour32 <- save_history(bbhM2, guided_tour(invConvexIndex()))
  bbhTourFull32 <- as.list(interpolate(bbhTour32))
  save(bbhTour32, bbhTourFull32, file = "cache/bbhGuidedConvex2.rda")
} else {
  load("cache/bbhGuidedConvex2.rda")
}

@@

#### Effect of random starts, and subsets used

The initial conditions for the optimisation, and the subset of variables used, can have a large effect on the projections returned. We illustrate this using only the splines2D index, and find that there is one more association that can be learned that was masked earlier.

Fig. \ref{fig:bbhGuided3} shows six maxima obtained by different starts, for two types of parameters: first, spin related parameters (i.e. alpha, theta\_jn, chi\_tot and chi\_p), and second position related parameters (i.e. ra, dec and distance). Four of the six (a-d) are almost identical, but not interesting projections. Projection f has the highest PP index value but it is primarily the view seen in the bivariate plot of dec and ra.

Choosing a different subset of variables reveals something new. The subspace of m1, ra, chi\_tot, alpha, distance, dec produces a more refined view of Fig. \ref{fig:bbhGuided3} projection f. When alpha contributes in contrast to dec, the relationship between the points is almost perfectly on a curve. This is shown in Figure \ref{constructedExample}. Manually reconstructing this can be done by differencing the two parameters, in their original units. \textcolor{red}{This result is unexpected, it can be understood from ????? (Does it mean that knowing the precision angle gives more precise information about the location? Knowing ra uniquely determines dec-alpha rather than dec it seems.)}

@@

<!--

@@

-->
